---
title: "FinalProject"
Name: "Preeti Sonke"\
output: html_document
---

```{r}
# Import needed packages
#library(tidyverse)
library(ggplot2) 
#library(tigerstats) 
#library(reticulate)
library(MASS)
library(MLmetrics)
library(dplyr)
```
## Part 1 : Data reading and Visualization

```{r setup, include=FALSE}
ins <- read.csv("C:\\Users\\sonke\\Documents\\Programing\\Final Project\\insurance.csv", header=TRUE, stringsAsFactors = FALSE) 
```

```{r}
head(ins)
```


#To retrive the structure of the data
```{r}
str(ins)
```
# Number of observations (no of rows)

```{r}
nrow(ins)

```
> This dataset has 348 rows

# number of attributes ( no of columns)

```{r}
ncol(ins)
```
>This dataset has 7 colums 


```{r}
dim(ins)
```
>dimesion of insurance dataset are 348 rows and 7 colums  

#Displying summary of dataset
```{r}
summary(ins)
```

# Showing Missing values

```{r}
nullValues<-sum(is.na(ins))
nullValues
```
# Finding duplicate values 
```{r}
sum(duplicated(ins))

```

>found 0 duplicate values 

```{r}
sapply(ins, class) 
```
> showing the datatype of each column variables .


#displaying the structure of the smoker column from the ins dataframe, 
```{r}
str(ins$smoker)
```


## Part 2: Graphical Representation

#Histogram distribution of insurance charges .

```{r}
hist(ins$charges, 
     main = "Distribution of Insurance Charges", 
     xlab = "Insurance Charges", 
     ylab = "Frequency", 
     col = "lightblue", 
     border = "black") 

```
>There's a higher frequency of insurance charges in the lower range (up to approximately $10,000).
The frequency decreases as the charge amount increases.
There are fewer instances of very high insurance charges (beyond $30,000).
The distribution appears right-skewed, indicating that higher insurance charges are less common.
The majority of insurance charges fall within a modest range, suggesting most insurance costs are not extremely high.

##Histogram of Insurance Charges with Q1 and Q3

```{r}


# To Calculate Q1 and Q3 for charges
Q1 <- quantile(ins$charges, 0.25)
Q3 <- quantile(ins$charges, 0.75)


hist(ins$charges, 
     main = "Histogram of Insurance Charges with Q1 and Q3", 
     xlab = "Insurance Charges", 
     ylab = "Frequency", 
     col = "lightblue", 
     border = "black")

# Add vertical lines for Q1 and Q3
abline(v = Q1, col = "black", lwd = 2, lty = 2)
abline(v = Q3, col = "black", lwd = 2, lty = 2)



```
>The histogram for insurance charges with marked Q1 and Q3 indicates that the middle 50% of the data (the interquartile range) lies between these two quartiles. The Q1 marker shows where 25% of the charges fall below, and the Q3 marker shows where 25% of the charges fall above, 

#Boxplot of Charges by Smoker Status
```{r}
# Assuming 'smoker' is a factor with two levels: 0 for non-smokers and 1 for smokers
boxplot(charges ~ smoker, data = ins,
        main = "Boxplot of Charges by Smoker Status",
        col = c("lightblue", "lightgreen"))  
```

>The boxplot illustrates that insurance charges are significantly higher for smokers compared to non-smokers, with smokers showing a wider range and higher median of charges. Non-smokers have a tighter distribution of charges with a lower median and fewer outliers.



#Overlay Histogram of Charges for Smokers vs. Non-Smokers

```{r}
library(ggplot2)
ins$smoker <- as.factor(ins$smoker)


ggplot(ins, aes(x = charges, fill = smoker)) +
  geom_histogram(aes(y = ..density..), position = 'identity', alpha = 0.5, bins = 30) +
  labs(title = 'Overlay Histogram of Charges for Smokers vs. Non-Smokers',
       x = 'Charges',
       y = 'Density') +
  scale_fill_manual(values = c('blue', 'red'))


```
>The histogram overlays the density of insurance charges for smokers (red) and non-smokers (blue).
Smokers tend to have higher insurance charges, as indicated by the red bars extending further along the x-axis.
Both groups show a right-skewed distribution, with the majority of charges clustered at the lower end.



## Part 3:Data Analysis 
```{r}
install.packages("Metrics")
library(Metrics)
```



##Hypothesis testing

```{r}
t.test(charges ~ smoker, data=ins)
```
>Hypothesis testing results for charges by smoker status:
Null Hypothesis: No difference in average charges between smokers and non-smokers
Alternative Hypothesis: Average charges differ between smokers and non-smokers
Test Used: Two sample t-test
Results:
t-statistic = -18.192, df = 94.439, extremely low p-value <0.0001
Average charges significantly higher for smokers (mean=$32,313.946) vs non-smokers (mean=$8,465.493)
95% CI excludes zero, indicating observed mean difference very unlikely to include true population difference of zero
Conclusion:
Very strong statistical evidence to reject the null hypothesis
Compelling support that smoker status has highly significant impact on charges
Alternative hypothesis supported - average charges differ significantly between smokers and non-smokers in the population

```{r}
t.test(charges ~ sex, data=ins)
```
> Hypothesis testing results for charges by sex:
Null Hypothesis: No difference in average charges between males and females
Alternative Hypothesis: Average charges differ between males and females
Test Used: Two sample t-test
Results:
t-statistic = -1.841, df = 335.22, p-value = 0.0665
Average charges higher for males (mean=$15,233.78) vs females (mean=$12,756.36)
95% CI includes zero, ranging from -$5,124 to $170
Conclusion:
P-value above the typical 0.05 threshold, so the null hypothesis cannot be rejected
Evidence is insufficient to determine a statistically significant difference in average charges by sex
While point estimates suggest higher charges for males, difference could also be due to chance
Alternative hypothesis of different average charges by sex is not strongly supported based on this sample data

Linear Regression

here we have take charges as response variable in linear regression as-----
Assumptions:

Null Hypothesis: There is no linear relationship between charges and the predictors age, sex, bmi, smoker, children, region
Alternative Hypothesis: There is a linear relationship between charges and the predictors
Performing F-test on the linear regression model to check whether group of predictor have impact on response variable or not.

```{r}
model <- lm(charges ~ age + sex + bmi + smoker + children + region, data = ins)
summary(model)
```
>The p-value of the F-statistic is very small (< 2.2e-16),so we can reject the null hypothesis at any reasonable significance level concluding that there is a linear relationship between charges and the predictors



Linear Regression with Subset Selection
Here is the model with all possible predictors

```{r}
full_model <- lm(charges ~ ., data = ins)
summary(full_model)
```
backward stepwise regression selection based on AIC

```{r}
back_final_model <- stepAIC(full_model, direction="backward")
```
```{r}
forward_final_model <- stepAIC(full_model, direction="forward")
```
>The final model contains significant predictors:
age, sex, bmi, smoker, children, region

To assess we choose backward final model

```{r}
summary(back_final_model)
```
>The model fits the data well, explaining around 77% of the variance in charges based on the high R-squared value.
The regression model captures most of the variation in charges using statistically significant predictor variables.
It fits the data very well and outperforms a simple intercept model.
Predictions have reasonably low error, suggesting the model predicts charges with acceptable accuracy.

Predict in test dataset

```{r}
test <- ins[sample(nrow(ins), 0.25*nrow(ins)),]
pred <- predict(back_final_model, newdata=test)
MAE(y_pred = pred,y_true = test$charges)
MSE(y_pred = pred,y_true = test$charges)
```

```{r}
par(mfrow=c(2,2))
plot(back_ffinal_model)
```

```{r}
init <- lm(charges~1, data=ins)
full <- lm(charges ~ ., data=ins)


forwardModel <- stepAIC(init, scope=formula(full), direction="forward")


final_formula <- formula(forwardModel)

test <- ins[sample(nrow(ins), 0.25),]
pred <- predict(forwardModel, newdata=test)
MAE(pred, test$charges)
MSE(pred, test$charges)


print(final_formula)

```

